\documentclass {article}

\title {
	Seminar: Advanced Topics in \\
	Quantum Computing \\
	On efficient encodings for quantum \\
	solutions to vehicle routing problems
}
\author {Eben Jowie Haezer}
\date {\today}

\begin {document}
\maketitle

\begin {abstract}
This report details recent advances in the optimisation of computing resources
for quantum approaches to solving the vehicle routing problem (VRP) and 
its variants. This set of problems is of significant importance with regard
to logistical applications in industry. In accordance with the input
constraints of the quantum hardware, the problem is formulated as a quadratic
unconstrained binary optimisation (QUBO). A simple approach known as the
full encoding results in each solution represented by a unique basis state,
thereby requiring one qubit per classical variable. Due to this inefficiency
in resource allocation when considering the at worst factorial search space,
a more optimised minimal encoding is suggested that offers a logarithmic
reduction in necessary computing power. In spite of certain drawbacks incurred
by employing this minimal encoding, experiments have shown that the solution
quality is not heavily impacted.
\end {abstract}

% structure:
%
% i.	introduction
% ii.	vrp
% iii.	qubo
% iv.	encoding schemes
% v.	discussion
% vi.	conclusion
% vii.	appendix

\section {Introduction}
The Vehicle Routing Problem (VRP) is concerned with finding optimal routes
for vehicles to deliver goods to a set of customers with some geographical
distance between them. It is obvious that this type of problem has far
reaching practical applications in various contexts, in fact the need to find
a reasonable solution is almost ubiquitous when dealing with logistical
planning, for example to determine efficient road, rail, shipping, and air
routes for commercial or public interest.

VRP, being itself a more general version of the Travelling Salesman Problem
(TSP), is similarly an NP hard combinatorial optimisation problem with a
solution space scaling factorially in the number of customers, and thus
finding the definitive optimal solution already becomes intractable at
two digit customer counts. The most optimal classical algorithms known to date
and used in practice employ heuristics and greedy methods to construct routes
within single digit percentage tolerances.

More recently, the noisy intermediate scale quantum (NISQ) era has paved
the way for the development of variational quantum algorithms alongside the
potent quantum annealing method as a contender for computing reliable and
near optimal solutions to combinatorial optimisation problems, exploiting
the unique unique inherent property of quantum algorithms to operate on the
entire solution space at once, albeit incrementally and probabilistically.

Initial tests have shown some promise in applying these quantum algorithms
to the VRP against classical solvers, achieving comparable accuracy on
small problem instances. As quantum hardware continues to improve, it makes
sense to also look at utilising it as efficiently as possible. Solving VRP
naively using quantum methods maps one classical variable to one whole qubit,
quickly exhausting the already limited computing resources when introducing
additional constraints such as vehicle capacity or time windows on larger
problem scales more commonly encountered in practical scenarios.

This report discusses recent research into an idea to mitigate this
uneconomical use of computing power through a more clever and refined 
encoding of the input problem, in order to achieve a logarithmic correlation
between problem size and resource consumption.
Section II outlines a formal description of the VRP and some of its variants.
Section III explains the QUBO and its application to solving the VRP on
quantum hardware, along with a brief mention of the closely related
Ising model and in particular its Hamiltonian function.
Section IV details the encoding approaches onto the quantum system and lists
benefits and drawbacks for each.
Published experimental observations describing the impact of the encodings
are discussed in Section V,
and the report concludes in Section VI.

\section {Vehicle Routing Problem}
The vehicle routing problem (VRP) is a generalisation of the perhaps more
well known NP hard Travelling Salesman Problem (TSP). VRP seeks the optimal
solution 

\section {Problem Formulation}
The quadratic unconstrained binary optimisation (QUBO) 

\section {Encoding Approaches}

\section {Discussion}

\section {Conclusion}

\section {Appendix}

\end {document}

